%%% PUBHBIO 6210 - PRACTICE EXERCISES
\documentclass[17pt,landscape]{foils}

% Include command file
\usepackage{commandsExercises}
 
 %print numbers
%\toggletrue{solutions}
\togglefalse{solutions}

% Exercises
\begin{document}

\exercise{4D}{A test of the null hypothesis $H_0: \mu=\mu_0$ gives test statistic $z=-2.12$.

(a) What is the p-value if the alternative is $H_a: \mu \ne \mu_0$?

\answer{4cm}{ANSWER: 0.0340\\
~\\
p-value = $P(Z > |-2.12|) + P(Z < -|-2.12|)$\\
~\quad\qquad$=P(Z>2.12)+P(Z<-2.12)$\\
~\quad\qquad$=2 \times P(Z>2.12)$ by symmetry\\
~\quad\qquad$= 2\times 0.0170 = 0.0340$
}

(b) What is the p-value if the alternative is $H_a: \mu < \mu_0$?

\answer{4cm}{ANSWER: 0.0170\\
~\\
p-value = $P(Z < -2.12)$\\
~\quad\qquad$=P(Z>2.12)$ by symmetry\\
~\quad\qquad$= 0.0170$}

(c) What is the p-value if the alternative is $H_a: \mu > \mu_0$?

\answer{2cm}{ANSWER: 0.983\\
~\\
p-value = $P(Z > -2.12)$\\
~\quad\qquad$=1 - P(Z<-2.12)$ by complement rule\\
~\quad\qquad$=1 - P(Z>2.12)$ by symmetry\\
~\quad\qquad$= 1-0.0170 = 0.983$}

}

%%%
\newpage

\exercise{4D}{The p-value for a two-sided significance test of $H_0: \mu=9$ is 0.082.

(a) Do you reject the null hypothesis at level $\alpha=0.05$?

\answer{3cm}{ANSWER: No, since 0.082 $>$ 0.05}

(b) Do you reject the null hypothesis at level $\alpha=0.1$?

\answer{3cm}{ANSWER: Yes, since 0.082 $<$ 0.1}

(c) Would a 95\% confidence interval for $\mu$ contain 9? Why?

\answer{3cm}{ANSWER: Yes. Since we fail to reject at $\alpha=0.05$, the null hypothesis value of 9 will be in the 95\% confidence interval.}

(d) Would a 90\% confidence interval for $\mu$ contain 9? Why?

\answer{1.5cm}{ANSWER: No. Since we reject at the $\alpha=0.1$ level, the null hypothesis value of 9 will not be in the 90\% confidence interval.}
}

%%%
\newpage

\exercise{4D}{For each of the following scenarios, state whether a type I error, a type II error, or no error will be made. Assume all tests are two-sided at the $\alpha = 0.05$ level.

(a) A test of $H_0: \mu = 10$ produces a p-value = 0.31 and the true population mean is 14.

\answer{4cm}{ANSWER: Type II error\\
~\\
The p-value 0.31 $>$ 0.05, so you fail to reject the null, but this is the wrong conclusion because the null hypothesis is not true (truth is $\mu$ = 14). Failing to reject when you should reject is a Type II error.}

(b) A test of $H_0: \mu=6$ produces a p-value = 0.02, and the true population mean is 10.

\answer{4cm}{ANSWER: No error made.\\
~\\
The p-value 0.02 $<$ 0.05, so you reject the null, which is the correct conclusion because the null is not true (truth is $\mu$ = 10).}

(c) A test of $H_0: \mu=12$ produces a p-value = 0.005 and the true population mean is 12.

\answer{4cm}{ANSWER: Type I error\\
~\\
The p-value 0.005 $<$ 0.05, so you reject the null, but this is the wrong conclusion because the null hypothesis is true (truth is $\mu$ = 12). Rejecting when you shouldn't reject is a Type I error.}

}

\end{document}
