---
title: "Standard Normal Distribution Table"
format:
  html:
    embed-resources: true
    toc: false
    html-math-method: mathjax
---

:::: {.columns}

::: {.column width = "40%"}

```{r}
#| echo: false
#| fig-cap: ""
#| fig-alt: "A standard normal distribution curve (bell curve) centered at zero. The region under the curve from the far left up to a positive value 'x' is shaded, illustrating the cumulative probability P(Z < x) mentioned in Column A of the table."

# Graph of area to the left
par(mar=c(2.1,0.3,0.3,0.3))
curve(dnorm(x),-4,4, lwd=2, lty=1, frame.plot=FALSE, yaxt="n", xaxt="n", ylim=c(0,0.45), yaxs="i")
axis(1, at=c(-5,0,1,5), labels=c(" ","0","x"," "), lwd=2, cex.axis=1.8)
# P(Z < x)
xvals <- seq(-4,1,length=50)
dvals <- dnorm(xvals)
polygon(c(xvals, rev(xvals)), c(rep(0,50),rev(dvals)), col="grey90")
text(-3,0.3,expression(P*(Z<x)), cex=1.8, pos=1)
arrows(-3, 0.22, -0.5, 0.15, code=2, lwd=2, length=0.1)
text(-4.2,0.42,expression(underline("Column A")),cex=1.5, pos=4)
```

:::

::: {.column width = "10%"}

:::

::: {.column width = "40%"}

```{r}
#| echo: false
#| fig-cap: ""
#| fig-alt: "A standard normal distribution curve (bell curve) centered at zero. The region under the curve from a positive value 'x' to the far right is shaded, illustrating the cumulative probability P(Z > x) mentioned in Column B of the table."

# Graph of area to the right
par(mar=c(2.1,0.3,0.3,0.3))
curve(dnorm(x),-4,4, lwd=2, lty=1, frame.plot=FALSE, yaxt="n", xaxt="n", ylim=c(0,0.45), yaxs="i")
axis(1, at=c(-5,0,1,5), labels=c(" ","0","x"," "), lwd=2, cex.axis=1.8)
# P(Z > x)
xvals <- seq(1,4,length=50)
dvals <- dnorm(xvals)
polygon(c(xvals, rev(xvals)), c(rep(0,50),rev(dvals)), col="grey90")
text(3,0.3,expression(P*(Z>x)), cex=1.8, pos=1)
arrows(3, 0.22, 1.3, 0.08, code=2, lwd=2, length=0.1)
text(4.2,0.42,expression(underline("Column B")),cex=1.5, pos=2)
```

:::

::::

Use this table to look up probabilities for the standard normal distribution. Locate your Z-score in the x column to find the corresponding area to the left (Column A) or the area to the right (Column B). The table is scrollable, and you can use the search function in your browser to jump to a specific value.

```{r}
#| echo: false
#| message: false
library(dplyr)
library(gt)

z_values <- seq(0.00, 3.77, by = 0.01)

z_df <- data.frame(
  x = z_values,
  lower = pnorm(z_values),               # P(Z < x)
  upper = pnorm(z_values, lower = FALSE) # P(Z > x)
)

z_df %>%
  gt() %>%
  tab_header(title = "Normal Distribution") %>%
  opt_css(css = "
    thead {
      position: sticky;
      top: 0;
      z-index: 100;
      background-color: white;
    }
    .gt_column_spanner_outer {
      position: sticky;
      top: 0;
    }
  ") %>%
  tab_options(
    container.height = px(500),
    container.overflow.y = "scroll"
  ) %>%
  tab_spanner(label = "A", columns = lower) %>%
  tab_spanner(label = "B", columns = upper) %>%
  cols_label(x = "x", lower = "P(Z < x)", upper = "P(Z > x)") %>%
  fmt_number(columns = lower:upper, decimals = 4) %>%
  cols_align(align = "center", columns = everything()) %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = list(
              cells_column_labels(),
              cells_column_spanners())) 
```

Note: Column A is the area to the left; Column B is the area to the right.
